Taking Linear Logic Apart

### 2018-04-28 doc/tlla2018/main.tex
Minor fixes
minor changes

\documentclass[draft,submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{Linearity and TLLA 2018}
\usepackage[british]{babel}
\usepackage[all]{foreign}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\allowdisplaybreaks
\usepackage[inline]{enumitem}
\usepackage{breakurl} % Not needed if you use pdflatex only.
\usepackage{underscore} % Only needed if you use pdflatex.
\usepackage{textcomp,textgreek,upgreek}
\usepackage{nameref,hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\hypersetup{hidelinks,final}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{example}{Example}
\input{preamble-typing}
\usepackage[sectionbib,square,numbers,sort&compress]{natbib}
\renewcommand*{\eg}{e.g.\@\xspace}
\renewcommand*{\ie}{i.e.\@\xspace}
\usepackage{more-space}
\usepackage{setspace}
\renewcommand{\baselinestretch}{0.925}

@@

Reductions relate processes with their reduced forms \eg, a reduction

@@

\paragraph{Types}
Channels in \cp are typed using a session type system which corresponds to
classical linear logic.
A process communicating over a channel of type \ty{A \tens B} must be ready to
split into two independent processes (two processes which share no
channels), ready to communicate over a channel of type \ty{A} and \ty{B},
respectively, and process communicating over a channel of type \ty{A \parr B} is
dual to this, and therefore must be ready to start communicating concurrently
with two independent processes.
A process communicating over a channel of type \ty{A \plus B} can choose to
communicate either as \ty{A} or \ty{B}, and a process communicating over a
channel of type \ty{A \with B} must be able to communicate either as \ty{A} or
\ty{B}.
\begin{definition}[Types]\label{def:cp-types}
  \[
    \begin{array}{rclrcl}
      \ty{A}, \ty{B}, \ty{C}
           :=& \ty{A \tens B} &\text{pair of independent processes}
      &  \mid& \ty{\one}      &\text{unit for} \; {\tens}
      \\ \mid& \ty{A \parr B} &\text{pair of interdependent processes}
      &  \mid& \ty{\bot}      &\text{unit for} \; {\parr}
      \\ \mid& \ty{A \plus B} &\text{internal choice}
      &  \mid& \ty{\nil}      &\text{unit for} \; {\plus}
      \\ \mid& \ty{A \with B} &\text{external choice}
      &  \mid& \ty{\top}      &\text{unit for} \; {\with}
    \end{array}
  \]
\end{definition}\noindent
Duality plays a crucial role in both linear logic and session types.
In \cp, the two endpoints of a channel are assigned dual types.
This ensures that, for instance, whenever a process \emph{sends} across a
channel, the process on the other end of that channel is waiting to
\emph{receive}.
Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$.
Duality is an involutive function.
\begin{definition}[Duality]\label{def:cp-negation}
  \[
    \setlength{\arraycolsep}{3pt}
    \begin{array}{lclclclclclclcl}
               \ty{(A \tens B)^\bot} & = & \ty{A^\bot \parr B^\bot}
      &\qquad& \ty{\one^\bot}        & = & \ty{\bot}
      &\qquad& \ty{(A \parr B)^\bot} & = & \ty{A^\bot \tens B^\bot}
      &\qquad& \ty{\bot^\bot}        & = & \ty{\one}
      \\       \ty{(A \plus B)^\bot} & = & \ty{A^\bot \with B^\bot}
      &\qquad& \ty{\nil^\bot}        & = & \ty{\top}
      &\qquad& \ty{(A \with B)^\bot} & = & \ty{A^\bot \plus B^\bot}
      &\qquad& \ty{\top^\bot}        & = & \ty{\nil}
    \end{array}
  \]
\end{definition}\noindent
Environments associate channels with types.
Names in environments must be unique, and environments $\ty{\Gamma}$ and
$\ty{\Delta}$ can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if
$\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} := \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}
  \) 
\end{definition}\noindent
\begin{figure}[htb]
  \begin{center}\cpInfAx \cpInfCut \end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfTens \cpInfParr \cpInfOne \end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfBot \cpInfPlus1 \cpInfPlus2\end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfWith \cpInfNil \cpInfTop \end{center}
  \caption{Classical Processes, typing.}
  \label{fig:cp}
\end{figure}\noindent
Typing judgements associate processes with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}}$ denotes
  that the process \tm{P} communicates along channels $\tm{x_1}\dots\tm{x_n}$
  following protocols $\ty{A_1}\dots\ty{A_n}$.
  Typing judgements are derived using rules in \cref{fig:cp}.
\end{definition}\noindent

@@

In this section, we introduce Hypersequent Classical Processes (\hcp), a variant
of \cp which registers parallelism in the typing jugdgements using
hypersequents, allowing us to take apart the monolithic term constructors of \cp
(\eg $\tm{\cpSend{x}{y}{P}{Q}}$) into the corresponding \textpi-calculus term
constructs.

@@

\paragraph{Types}
We use the same definitions for types and environments for \hcp as we used for \cp.
However, we introduce a new layer on top of sequents: hypersequents.
As \cp is a one-sided logic, and it uses the left-hand side of the
turnstile to write the process, the traditional hypersequent notation can look
confusing: ``$\seq[P]{\Gamma_1}\hsep\dots\hsep\seq{\Gamma_n}$'' seems to claim
that $\tm{P}$ acts according to protocol $\ty{\Gamma_1}$.
What are all the other $\Gamma$s doing there? Are they typing empty processes?
Therefore, we opt to leave out the repeated turnstile, and instead work with the
notion of ``hyper-environments''. However, we will still refer to our system as a
hypersequent system.
Hyper-environments are either empty, or consist of a series of typing
environments, separated by vertical bars.
A hyper-environment $\ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ types a series of
$n$ entangled, but independent processes.
\begin{definition}[Hyper-environment]\label{def:hcp-hyper-environment}
  \(
  \ty{\mathcal{G}}, \ty{\mathcal{H}} :=
  \ty{\emptyhypercontext}
  \; \mid \;
  \ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}
  \)
\end{definition}\noindent
Typing judgements in \hcp associate processes with hyper-environments.
\textsc{H-Mix} composes two processes in parallel, but remembers that they are
independent on in the sequent.
\textsc{H-Cut} and $(\tens)$ take as their premise a process which consists of
at least two unconnected processes, and connects them, eliminating the vertical
bar.
Note that each logical rule has the side condition that $\tm{x} \not\in
\mathcal{G}$, which can be read as ``you cannot act on one end-point of $x$ if
you are also holding its other end-point''.
This prevents self-locking processes, \eg
$\tm{\piBoundSend{x}{}{\piRecv{x}{}{\piHalt}}}$.
\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes
  that the process $\tm{P}$ consists of $n$ independent, but potentially
  entangled processes, each of which communicates according to its own
  protocol $\Gamma_i$.
  Typing judgements can be constructed using the inference rules in \cref{fig:hcp}.
  Furthermore, each logical rule has the side condition that $\tm{x} \not\in
  \ty{\mathcal{G}}$.
\end{definition}\noindent
\begin{figure}[htb]
  \begin{center}\hccpInfAx \hccpInfCycle\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfMix \hccpInfHalt\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBoundTens \hccpInfParr \hccpInfOne\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBot \hccpInfPlus1 \hccpInfPlus2\end{center}
  \vspace*{0.5\baselineskip}
  \begin{center}\hccpInfWith \hccpInfNil \hccpInfTop\end{center}
  \caption{Hypersequent Classical Processes, typing.}
  \label{fig:hcp}
\end{figure}

@@

We start out by showing that we can interpret environments as types by
collapsing them using a series of pars.
\begin{definition}\label{def:bigparr}
  \(
  \ty{\bigparr(\tmty{x_1}{A_1} , \dots , \tmty{x_n}{A_n})} =
  \ty{A_1 \parr \dots \parr A_n}
  \)
\end{definition}\noindent
\begin{lemma}\label{lem:cp-bigparr}
  If $\seq{\Gamma}$ in \cp, then $\seq{\bigparr\Gamma}$ in \cp.
\end{lemma}\vspace*{-0.75\baselineskip}%
\begin{proof}
  By induction on the structure of the environment $\Gamma$.
  If $\ty{\Gamma} = \ty{A}$, then the statement is trivially true.
  If $\ty{\Gamma} = \ty{A, B, \Gamma'}$, then we apply $(\parr)$ to get $\seq{A
    \parr B, \Gamma'}$, and apply the induction hypothesis.
\end{proof}\noindent
Then, using \cref{lem:hcp-disentangle}, we can show that we can interpret
hyper-environments as types by collapsing them using a series of tensors.
In the case of the empty hyper-environment $\ty{\emptyhypercontext}$, we
interpret this as the unit of tensor.
\begin{definition}\label{def:bigtens}
  \(
  \begin{array}{lcl}
    \ty{\bigtens(\emptyhypercontext)}
    & = & \ty{\one}
    \\
    \ty{\bigtens(\Gamma_1 \hsep \dots \hsep \Gamma_n)}
    & = & \ty{\bigparr\Gamma_1 \tens \dots \tens \bigparr\Gamma_n}
  \end{array}
  \)
\end{definition}\noindent

@@

\looseness=-1
The idea of using hypersequents to capture parallelism in linear logic judgements is not
novel: \citet{carbone2018} extended the multiplicative-additive fragment
of intuitionistic linear logic with hypersequents to type global descriptions of
process communications known as choreographies.
This work is distinct from our approach in that \hcp is based on classical linear
logic and manipulates hypersequents differently: in \citet{carbone2018}, hypersequents can be formed only when sequents share
resources (\cf \textsc{H-Mix}), and resource sharing is then explicitly tracked
using an additional connection modality (which is not present in \hcp).

>>>>

\documentclass[draft,submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{Linearity and TLLA 2018}
\usepackage[british]{babel}
%\usepackage[all]{foreign}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\allowdisplaybreaks
\usepackage[inline]{enumitem}
\usepackage{breakurl} % Not needed if you use pdflatex only.
\usepackage{underscore} % Only needed if you use pdflatex.
\usepackage{textcomp,textgreek,upgreek}
\usepackage{nameref,hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\hypersetup{hidelinks,final}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{example}{Example}
\input{preamble-typing}
\usepackage[sectionbib,square,numbers,sort&compress]{natbib}
%\renewcommand*{\eg}{e.g.\@\xspace}
%\renewcommand*{\ie}{i.e.\@\xspace}
\usepackage{more-space}
\usepackage{setspace}
\renewcommand{\baselinestretch}{0.925}

@@

Reductions relate processes with their reduced forms e.g., a reduction

@@

\paragraph{Types}
Channels in \cp are typed using a session type system which corresponds to
classical linear logic.
A process communicating over a channel of type \ty{A \tens B} must be ready to
split into two independent processes (two processes which share no
channels), ready to communicate over a channel of type \ty{A} and \ty{B},
respectively, and process communicating over a channel of type \ty{A \parr B} is
dual to this, and therefore must be ready to start communicating concurrently
with two independent processes.
A process communicating over a channel of type \ty{A \plus B} can choose to
communicate either as \ty{A} or \ty{B}, and a process communicating over a
channel of type \ty{A \with B} must be able to communicate either as \ty{A} or
\ty{B}.
\begin{definition}[Types]\label{def:cp-types}
    \begin{array}{rclrcl}
      \ty{A}, \ty{B}, \ty{C}
           :=& \ty{A \tens B} &\text{pair of independent processes}
      &  \mid& \ty{\one}      &\text{unit for} \; {\tens}
      \\ \mid& \ty{A \parr B} &\text{pair of interdependent processes}
      &  \mid& \ty{\bot}      &\text{unit for} \; {\parr}
      \\ \mid& \ty{A \plus B} &\text{internal choice}
      &  \mid& \ty{\nil}      &\text{unit for} \; {\plus}
      \\ \mid& \ty{A \with B} &\text{external choice}
      &  \mid& \ty{\top}      &\text{unit for} \; {\with}
Duality plays a crucial role in both linear logic and session types.
In \cp, the two endpoints of a channel are assigned dual types.
This ensures that, for instance, whenever a process \emph{sends} across a
channel, the process on the other end of that channel is waiting to
\emph{receive}.
Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$.
Duality is an involutive function.
\begin{definition}[Duality]\label{def:cp-negation}
    \begin{array}{lclclclclclclcl}
               \ty{(A \tens B)^\bot} & = & \ty{A^\bot \parr B^\bot}
      &\qquad& \ty{\one^\bot}        & = & \ty{\bot}
      &\qquad& \ty{(A \parr B)^\bot} & = & \ty{A^\bot \tens B^\bot}
      &\qquad& \ty{\bot^\bot}        & = & \ty{\one}
      \\       \ty{(A \plus B)^\bot} & = & \ty{A^\bot \with B^\bot}
      &\qquad& \ty{\nil^\bot}        & = & \ty{\top}
      &\qquad& \ty{(A \with B)^\bot} & = & \ty{A^\bot \plus B^\bot}
      &\qquad& \ty{\top^\bot}        & = & \ty{\nil}
Environments associate channels with types.
Names in environments must be unique, and environments $\ty{\Gamma}$ and
$\ty{\Delta}$ can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if
$\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} := \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}
  \) 
%\begin{figure}[htb]
%  \begin{center}\cpInfAx \cpInfCut \end{center}
%  \vspace*{-.5\baselineskip}
%  \begin{center}\cpInfTens \cpInfParr \cpInfOne \end{center}
%  \vspace*{-.5\baselineskip}
%  \begin{center}\cpInfBot \cpInfPlus1 \cpInfPlus2\end{center}
%  \vspace*{-.5\baselineskip}
%  \begin{center}\cpInfWith \cpInfNil \cpInfTop \end{center}
%  \caption{Classical Processes, typing.}
%  \label{fig:cp}
%\end{figure}\noindent
Typing judgements associate processes with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}}$ denotes
  that the process \tm{P} communicates along channels $\tm{x_1}\dots\tm{x_n}$
  following protocols $\ty{A_1}\dots\ty{A_n}$.
  Typing judgements are derived using the rules below.%rules in \cref{fig:cp}.
  \begin{center}\cpInfAx \cpInfCut \end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfTens \cpInfParr \cpInfOne \end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfBot \cpInfPlus1 \cpInfPlus2\end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfWith \cpInfNil \cpInfTop \end{center}

@@

In this section, we introduce Hypersequent Classical Processes (\hcp), a variant
of \cp which registers parallelism in the typing jugdgements using
hypersequents, allowing us to take apart the monolithic term constructors of \cp
(e.g.~$\tm{\cpSend{x}{y}{P}{Q}}$) into the corresponding \textpi-calculus term
constructs.

@@

\paragraph{Types}
We use the same definitions for types and environments for \hcp as we used for \cp.
However, we introduce a new layer on top of sequents: hypersequents.
As \cp is a one-sided logic, and it uses the left-hand side of the
turnstile to write the process, the traditional hypersequent notation can look
confusing: ``$\seq[P]{\Gamma_1}\hsep\dots\hsep\seq{\Gamma_n}$'' seems to claim
that $\tm{P}$ acts according to protocol $\ty{\Gamma_1}$.
What are all the other $\Gamma$s doing there? Are they typing empty processes?
Therefore, we opt to leave out the repeated turnstile, and instead work with the
notion of ``hyper-environments''. However, we will still refer to our system as a
hypersequent system.
Hyper-environments are either empty, or consist of a series of typing
environments, separated by vertical bars.
A hyper-environment $\ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ types a series of
$n$ entangled, but independent processes.
\begin{definition}[Hyper-environment]\label{def:hcp-hyper-environment}
  \(
  \ty{\mathcal{G}}, \ty{\mathcal{H}} :=
  \ty{\emptyhypercontext}
  \; \mid \;
  \ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}
  \)
\end{definition}\noindent
Typing judgements in \hcp associate processes with hyper-environments.
\textsc{H-Mix} composes two processes in parallel, but remembers that they are
independent on in the sequent.
\textsc{H-Cut} and $(\tens)$ take as their premise a process which consists of
at least two unconnected processes, and connects them, eliminating the vertical
bar.
Note that each logical rule has the side condition that $\tm{x} \not\in
\mathcal{G}$, which can be read as ``you cannot act on one end-point of $x$ if
you are also holding its other end-point''.
This prevents self-locking processes, e.g.~$\tm{\piBoundSend{x}{}{\piRecv{x}{}{\piHalt}}}$.
\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes
  that the process $\tm{P}$ consists of $n$ independent, but potentially
  entangled processes, each of which communicates according to its own
  protocol $\Gamma_i$.
  Typing judgements can be constructed using the inference rules in \cref{fig:hcp}.
  Furthermore, each logical rule has the side condition that $\tm{x} \not\in
  \ty{\mathcal{G}}$.
\end{definition}\noindent
\begin{figure}[htb]
  \begin{center}\hccpInfAx \hccpInfCycle\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfMix \hccpInfHalt\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBoundTens \hccpInfParr \hccpInfOne\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBot \hccpInfPlus1 \hccpInfPlus2\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfWith \hccpInfNil \hccpInfTop\end{center}
  \caption{Hypersequent Classical Processes, typing.}
  \label{fig:hcp}
\end{figure}

@@

Environments can be interpreted as types by
collapsing them as series of pars.
\begin{definition}\label{def:bigparr}
  \(
  \ty{\bigparr(\tmty{x_1}{A_1} , \dots , \tmty{x_n}{A_n})} =
  \ty{A_1 \parr \dots \parr A_n}
  \)
\end{definition}\noindent
\begin{lemma}\label{lem:cp-bigparr}
  If $\seq{\Gamma}$ in \cp, then $\seq{\bigparr\Gamma}$ in \cp.
\end{lemma}\vspace*{-0.75\baselineskip}%
\begin{proof}
  By induction on the structure of the environment $\Gamma$.
  If $\ty{\Gamma} = \ty{A}$, then the statement is trivially true.
  If $\ty{\Gamma} = \ty{A, B, \Gamma'}$, then we apply $(\parr)$ to get $\seq{A
    \parr B, \Gamma'}$, and apply the induction hypothesis.
\end{proof}\noindent
Furthermore, hyper-environments can be interpreted as types by collapsing them as series of tensors where environments are handled using \cref{lem:hcp-disentangle}. The empty hyper-environment $\ty{\emptyhypercontext}$ (that is the unit of $\hsep$) is interpreted as the unit of tensor.
\begin{definition}\label{def:bigtens}
  \(
  \begin{array}{lcl}
    \ty{\bigtens(\emptyhypercontext)}
    & = & \ty{\one}
    \\
    \ty{\bigtens(\Gamma_1 \hsep \dots \hsep \Gamma_n)}
    & = & \ty{\bigparr\Gamma_1 \tens \dots \tens \bigparr\Gamma_n}
  \end{array}
  \)
\end{definition}\noindent

@@

\looseness=-1
The idea of using hypersequents to capture parallelism in linear logic judgements is not novel: \citet{carbone2018} extended the multiplicative-additive fragment of intuitionistic linear logic with hypersequents to type global descriptions of process communications known as choreographies.
This work is distinct from our approach in that \hcp is based on classical linear logic and manipulates hypersequents differently: in \citet{carbone2018}, hypersequents can be formed only when sequents share resources (cf.~\textsc{H-Mix}), and resource sharing is explicitly tracked
using an additional connection modality (which is not present in \hcp).
